A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2016-08-20 imaging_demo/ms.tex
Mark changes with boldface
From 1d980f4 (mnras):
====================

In ms.tex,

(*) Use boldface to mark changes compared to previous submitted
version

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. It takes advantage of the multiplication-convolution theorem of Fourier transforms. Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregular layouts and heterogeneous arrays of antennas. EPIC is highly modular and parallelizable, implemented in object oriented Python, and publicly available. We have verified the images produced to be equivalent to those produced using traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA). We have demonstrated its versatility to image with heterogeneous arrays by showing that it robustly estimates the input sky model while wrong assumptions about array homogeneity lead to systematic mis-estimates. Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying EPIC. Inherent availability of calibrated time-domain images on digitizer writeout time-scales and vastly lower I/O bandwidth relative to visibility-based systems will make it a prime candidate for transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena. 
\end{abstract}

@@

Radio astronomy is entering an era in which interferometers of hundreds to thousands of individual antennas are needed to achieve desired survey speeds. Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study of the history of hydrogen gas throughout the universe's evolution is pushing technology development towards arrays of low-cost antennas with large fields of view and densely packed layouts. Similarly, the search for transient objects and regular monitoring of the time-dependent sky is driving instruments in the same direction with the added requirement of fast read-outs. A number of new telescopes are being or were developed around the world based on this new paradigm, including the Hydrogen Epoch of Reionization Array\footnote{http://reionization.org} (HERA; \citealt{deb16}), the Murchison Widefield Array (MWA; \citealt{tin13,bow13}), the Donald~C.~Backer Precision Array for Probing the Epoch of Reionization (PAPER; \citealt{par10}), the LOw Frequency ARray (LOFAR; \citealt{van13}), the Canadian Hydrogen Intensity Mapping Experiment (CHIME; \citealt{ban14}), the Long Wavelength Array (LWA; \citealt{ell13}), and the low frequency Square Kilometer Array (SKA1-Low; \citealt{mel13}).

@@

We begin with a technical description of the algorithm in \S\ref{sec:math}, then discuss our particular implementation in \S\ref{sec:software}. We then verify the output data quality from our code in \S\ref{sec:verify} by presenting simulated images from both the EPIC correlator and comparing to a simulated FX correlator. We also demonstrate the performance with real-world data from the LWA. \S\ref{sec:versatility} demonstrates its ability to robustly image data from heterogeneous arrays. In \S\ref{sec:analysis}, we explore the scalability of the algorithm in the context of several array design choices. We identify specific array design classes where the EPIC correlator -- is computationally more efficient; and in the field of transients, demands significantly lesser I/O bandwidth relative to visibility-based approaches. We conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{2\pi i f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \nonumber\\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-2\pi i f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. The noise-like cross-terms ideally have zero mean. This expression can be further simplified using the sky brightness, $\mathcal{I}(\hat{\mathbf{s}},f,t)$, as $\left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t = \mathcal{I}(\hat{\mathbf{s}},f,t)\,\delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{B}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power pattern, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}\label{eqn:measurement-eqn-1}
  V_{ab}(f,t) &= \int \mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to {\bf the measurement plane}, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

@@

Using the optimal map-making formalism \citep{teg97a,teg97b}, a software holography image is formed using \citep{mor09}:
\begin{align}
  \boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}}) &= \mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{u})\,\mathbf{B}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v},\mathbf{v})\,\mathbf{m}(\mathbf{v}) \label{eqn:dirty-image-FX}
\end{align}
where the measured visibilities are weighted by the inverse of the system noise, followed by a gridding process using the holographic antenna power response as the gridding kernel, followed by a Fourier transform to create an image $\boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}})$. This is the optimal estimate of the true image $\boldsymbol{\mathcal{I}}(\hat{\mathbf{s}})$ given the visibility measurements.

@@

In a companion paper \citep{bea16}, we demonstrate a novel calibration 
technique (EPICal) which leverages the data products formed by direct imaging 
correlators to estimate antenna complex gains. This method correlates the antenna 
electric field signals with an image pixel from the output of the correlator in 
the feedback calibration fashion outlined in \citealt{mor11} (illustrated in 
Fig.~\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the 
calibration block). Furthermore it allows for arbitrarily complex sky models, 
and following the MOFF algorithm places no restriction on array layout, and 
accounts for non-identical antenna beam patterns. Direction dependent calibration 
can be achieved by correlating antenna signals with output pixels in the 
direction of $N_\textrm{c}$ calibration sources, then fitting for a functional 
model of the sky. Since antennas are only correlated with calibrator pixels, the 
computational complexity scales as $\sim N_\textrm{a} N_\textrm{c}$.

@@

Fig.~\ref{fig:MOFF-FX-image} shows the weighted dirty images (top) and synthesized beams (bottom) obtained with antenna-based MOFF and FX visibility-based imaging algorithms packaged in EPIC. The antenna auto-correlations that correspond to zero spacing have been removed from the correlated weights and data in the $uv$ plane, the MOFF image and the corresponding synthesized beam as described in \S\ref{sec:rm-autocorr}. The reconstructed sky image has the simulated sources at the expected sky positions in either case. Both algorithms result in images and synthesized beams that are well matched with each other. As expected, their fluxes are attenuated by a weighting proportional to the square of the antenna power pattern corresponding to a uniform square aperture.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure5}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed.eps}
  \caption{Weighted dirty images (top) and synthesized beams (bottom) obtained from simulated data using EPIC implementation of antenna-based MOFF algorithm (left) and visibility-based imaging (right). The antenna auto-correlations at zero-spacing have been removed from the MOFF images. The images in either case reconstruct the sources at the right locations with the fluxes attenuated expected after multiplication by the antenna power pattern. 
    The synthesized beams from the two algorithms are well matched in size and
    shape. The overall modulation by the power pattern is seen clearly in both
    images.}
  \label{fig:MOFF-FX-image}
\end{figure}

We examine in detail the respective synthesized beams in each case in Fig.~\ref{fig:MOFF-FX-psf-diff}. Slices at $m=0$ of the synthesized beams weighted by the antenna power pattern are shown for MOFF method using EPIC (solid black) and visibility-based (dashed gray) imaging. The two are found to match well. A magnified view shows that some differences at the level of $<0.5$\% exist in some regions. These are attributed to differences in the $uv$-plane antenna cross-correlation weights in the two methods which in turn arises due to the amount of coarseness in grid spacing as described below.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure6}
  \caption{Synthesized beams weighted by antenna power pattern sliced at $m=0$ obtained with MOFF algorithm using EPIC (solid black line) and visibility-based imaging (dashed gray line). The two appear almost identical. The inset provides a magnified view of differences in the synthesized beam slices between the two techniques at levels $\lesssim 0.5$\% relative to the peak. These are attributed to differences that arise in gridding and depends on the coarseness of the grid.}
  \label{fig:MOFF-FX-psf-diff}
\end{figure}

The left panel of Fig.~\ref{fig:MOFF-FX-uvwts-diff} shows differences (in percentage relative to peak) in cross-correlated weights obtained with MOFF imaging in EPIC and visibility-based imaging. The maximum difference appears near the center of the $uv$-plane corresponding to antenna auto-correlations which have not been perfectly removed in the former. In other regions of the $uv$-plane, the differences are of the order of a few percent. The reason for these differences is described below.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure7}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed.eps}
  \caption{{\it Left}: Differences (in \%) in cross-correlated weights in the $uv$ plane relative to the peak. The biggest difference ($\sim 20$\%) is found around the zero-spacing corresponding to auto-correlations of antenna weights due to gridding differences augmented by the relatively high number density of antenna auto-correlation footprints in that region. In most of the other regions, differences of the order of less than a few percent are seen. {\it Right}: Histogram (expressed as percentage) of the relative differences (in \%) between antenna cross-correlation weights shown on the left binned in intervals of 0.5\%. Over 70\% of the $uv$-cells do not differ by more than 0.5\% and over 90\% of $uv$-cells only differ by $<5$\% for the gridding coarseness used.} 
  \label{fig:MOFF-FX-uvwts-diff}
\end{figure}

@@

We study the effect of the differences in gridded weights on the image plane. Fig.~\ref{fig:image-psf-diff} shows the difference between the synthesized beams obtained with the two methods. A difference map between the two synthesized beams is shown on the left. The amplitude of the difference appears to be modulated by the directional power response of the antenna. On the right, in radial bins, the rms of the synthesized beam (gray) and the rms of the difference map (black) are plotted in percentage units relative to the peak (to be read using the axis on the left side of the plot). The antenna power pattern (red; to be read using the scale on the right) is plotted for reference.

@@

The synthesized beam rms is proportional to the antenna power pattern as 
expected from a point spread function uncorrected for the antenna power pattern. 
The rms of differenced synthesized beams is also modulated by the antenna power 
pattern. The rms of the difference is definitely lesser than the rms of the 
synthesized beam in the central regions up to $(l^2+m^2)^{1/2}\lesssim 0.3$. This 
implies that the beams are well matched in the central regions. In the outer 
regions, their mismatch is comparable to the rms of synthesized beams. These findings are consistent with Fig.~\ref{fig:MOFF-FX-psf-diff}. This 
indicates the two synthesized beams are not randomly different from 
each other in which case the rms of the difference would have been 
$\approx \sqrt{2}$ higher than the rms of the each of the synthesized beams. 
This means that while differences exist, large fractions of them are still 
well matched to each other even out to the horizon. Thus the rounding errors in 
gridding do not affect the statistics of the images or the synthesized beams.

@@

One of the advanced features of the MOFF algorithm and its implementation via EPIC is the ability to naturally account for heterogeneity of antennas while still producing images with maximal information \citep{mor11}. Unlike the assumption frequently made in standard radio interferometry that all antennas are identical, EPIC can implement the most generic case where no such assumption can be made. In other words, EPIC does not assume $W_a(\mathbf{r}-\mathbf{r}_a)$ or $B_{ab}(\mathbf{r}-\mathbf{r}_{ab})$ for different antennas are identical. However, any {\it a priori} knowledge available that certain sets of antennas have identical illumination patterns can be easily passed on to EPIC to avoid redundant computations and thus increase efficiency.

@@

This demonstrates that images from EPIC can robustly image data from heterogeneous arrays while being mathematically optimal. Besides being sub-optimal, erroneous assumptions about the homogeneity of an array can lead to significant and systematic mis-estimates (in this case, an underestimate) of the sky model.

@@

We now consider antenna array layouts described by three essential quantities in radio interferometry, namely, maximum baseline length, number of antennas, and the size of each antenna. Fig.~\ref{fig:parameter-space-bll-nant-instruments} shows the boundaries where the ratio of the number of computations required with visibility-based imaging relative to MOFF imaging is unity. The different colored lines correspond to different antenna sizes (cyan - 1~m$^2$, blue - 7~m$^2$, purple - 16~m$^2$, green - 28~m$^2$, orange - 150~m$^2$, red - 740~m$^2$, gray - 5900~m$^2$). Dashed lines of each color denote the boundary to the left of which the MOFF algorithm is favoured for the corresponding antenna size and vice versa for visibility-based imaging.

>>>>

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. It takes advantage of the multiplication-convolution theorem of Fourier transforms. Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregular layouts and heterogeneous arrays of antennas. EPIC is highly modular and parallelizable, implemented in object oriented Python, and publicly available. We have verified the images produced to be equivalent to those produced using traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA). {\bf We have demonstrated its versatility to image with heterogeneous arrays by showing that it robustly estimates the input sky model while wrong assumptions about array homogeneity lead to systematic mis-estimates.} Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying EPIC. Inherent availability of calibrated time-domain images on digitizer writeout time-scales and vastly lower I/O bandwidth relative to visibility-based systems will make it a prime candidate for transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena. 
\end{abstract}

@@

Radio astronomy is entering an era in which interferometers of hundreds to thousands of individual antennas are needed to achieve desired survey speeds. Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study of the history of hydrogen gas throughout the universe's evolution is pushing technology development towards arrays of low-cost antennas with large fields of view and densely packed layouts. Similarly, the search for transient objects and regular monitoring of the time-dependent sky is driving instruments in the same direction with the added requirement of fast read-outs. A number of new telescopes are being or were developed around the world based on this new paradigm, including the Hydrogen Epoch of Reionization Array\footnote{http://reionization.org} (HERA; \citealt{deb16}), the Murchison Widefield Array (MWA; \citealt{tin13,bow13}), the {\bf Donald~C.~Backer} Precision Array for Probing the Epoch of Reionization (PAPER; \citealt{par10}), the LOw Frequency ARray (LOFAR; \citealt{van13}), the Canadian Hydrogen Intensity Mapping Experiment (CHIME; \citealt{ban14}), the Long Wavelength Array (LWA; \citealt{ell13}), and the low frequency Square Kilometer Array (SKA1-Low; \citealt{mel13}).

@@

We begin with a technical description of the algorithm in \S\ref{sec:math}, then discuss our particular implementation in \S\ref{sec:software}. We then verify the output data quality from our code in \S\ref{sec:verify} by presenting simulated images from both the EPIC correlator and comparing to a simulated FX correlator. We also demonstrate the performance with real-world data from the LWA. {\bf \S\ref{sec:versatility} demonstrates its ability to robustly image data from heterogeneous arrays.} In \S\ref{sec:analysis}, we explore the scalability of the algorithm in the context of several array design choices. We identify specific array design classes where the EPIC correlator -- is computationally more efficient; and in the field of transients, demands significantly lesser I/O bandwidth relative to visibility-based approaches. We conclude and discuss future research prospects in \S\ref{sec:conclusions}.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{2\pi i f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \nonumber\\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-2\pi i f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. {\bf The noise-like cross-terms ideally have zero mean. This expression can be further simplified using the sky brightness}, $\mathcal{I}(\hat{\mathbf{s}},f,t)$, as $\left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t = \mathcal{I}(\hat{\mathbf{s}},f,t)\,\delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{B}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power pattern, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}\label{eqn:measurement-eqn-1}
  V_{ab}(f,t) &= \int \mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{B}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to {\bf the measurement plane}, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

@@

Using the optimal map-making formalism \citep{teg97a,teg97b}, a software holography image is formed using \citep{mor09}:
\begin{align}
  \boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}}) &= \mathbf{F}^\textrm{T}(\hat{\mathbf{s}},\mathbf{u})\,\mathbf{B}^{\,\textrm{T}}(\mathbf{u},\mathbf{v})\,\mathbf{N}^{-1}(\mathbf{v},\mathbf{v})\,\mathbf{m}(\mathbf{v}) \label{eqn:dirty-image-FX}
\end{align}
where the measured visibilities are weighted by the inverse of the system noise, followed by a gridding process using the holographic antenna power response as the gridding kernel, followed by a Fourier transform to create an image $\boldsymbol{\mathcal{I}}^\prime(\hat{\mathbf{s}})$. This is the {\bf optimally weighted} estimate of the true image $\boldsymbol{\mathcal{I}}(\hat{\mathbf{s}})$ given the visibility measurements.

@@

In a companion paper \citep{bea16}, we demonstrate a novel calibration 
technique (EPICal) which leverages the data products formed by direct imaging 
correlators to estimate antenna complex gains. This method correlates the antenna 
electric field signals with an image pixel from the output of the correlator in 
the feedback calibration fashion outlined in \citealt{mor11} (illustrated in 
Fig.~\ref{fig:MOFF-flowchart} by the arrow leading from the imager to the 
calibration block). Furthermore it allows for arbitrarily complex sky models, 
and following the MOFF algorithm places no restriction on array layout, and 
accounts for non-identical antenna beam patterns. Direction dependent calibration 
can be achieved by correlating antenna signals with output pixels in the 
direction of $N_\textrm{c}$ calibration sources, then fitting for a functional 
model of the sky. Since antennas are only correlated with calibrator pixels, the 
computational complexity scales as $\sim N_\textrm{a} N_\textrm{c}$. {\bf The calibration performance of EPIC is explored for varying levels of completeness of sky model in \citet{bea16}.}

@@

Fig.~\ref{fig:MOFF-FX-image} shows the weighted dirty images (top) and synthesized beams (bottom) obtained with antenna-based MOFF and FX visibility-based imaging algorithms packaged in EPIC. The antenna auto-correlations that correspond to zero spacing have been removed from the correlated weights and data in the $uv$ plane, the MOFF image and the corresponding synthesized beam as described in \S\ref{sec:rm-autocorr}. The reconstructed sky image has the simulated sources at the expected sky positions in either case. Both algorithms result in images and synthesized beams that are well matched with each other. {\bf As expected, their fluxes are attenuated by a weighting proportional to the square of the antenna power pattern corresponding to a uniform square aperture.}

\begin{figure}
  \includegraphics[width=\columnwidth]{figure5}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed.eps}
  \caption{Weighted dirty images (top) and synthesized beams (bottom) obtained from simulated data using EPIC implementation of antenna-based MOFF algorithm (left) and visibility-based imaging (right). The antenna auto-correlations at zero-spacing have been removed from the MOFF images. The images in either case reconstruct the sources at the right locations with the fluxes attenuated expected after multiplication by the antenna power pattern {\bf squared}. The synthesized beams from the two algorithms are well matched in size and shape. The overall modulation by the power pattern is seen clearly in both images.}
  \label{fig:MOFF-FX-image}
\end{figure}

{\bf We examine in detail the respective synthesized beams in each case in Fig.~\ref{fig:MOFF-FX-psf-diff}. Slices at $m=0$ of the synthesized beams weighted by the antenna power pattern are shown for MOFF method using EPIC (solid black) and visibility-based (dashed gray) imaging. The two are found to match well. A magnified view shows that some differences at the level of $<0.5$\% exist in some regions. These are attributed to differences in the $uv$-plane antenna cross-correlation weights in the two methods which in turn arises due to the amount of coarseness in grid spacing as described below.}

\begin{figure}
  \includegraphics[width=\columnwidth]{figure6}
  \caption{{\bf Synthesized beams weighted by antenna power pattern sliced at $m=0$ obtained with MOFF algorithm using EPIC (solid black line) and visibility-based imaging (dashed gray line). The two appear almost identical. The inset provides a magnified view of differences in the synthesized beam slices between the two techniques at levels $\lesssim 0.5$\% relative to the peak. These are attributed to differences that arise in gridding and depends on the coarseness of the grid.}}
  \label{fig:MOFF-FX-psf-diff}
\end{figure}

{\bf The left panel of Fig.~\ref{fig:MOFF-FX-uvwts-diff} shows differences (in percentage relative to peak) in cross-correlated weights obtained with MOFF imaging in EPIC and visibility-based imaging. The maximum difference appears near the center of the $uv$-plane corresponding to antenna auto-correlations which have not been perfectly removed in the former. In other regions of the $uv$-plane, the differences are of the order of a few percent. The reason for these differences is described below.}

\begin{figure}
  \includegraphics[width=\columnwidth]{figure7}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed.eps}
  \caption{{\bf {\it Left}: Differences (in \%) in cross-correlated weights in the $uv$ plane relative to the peak. The biggest difference ($\sim 20$\%) is found around the zero-spacing corresponding to auto-correlations of antenna weights due to gridding differences augmented by the relatively high number density of antenna auto-correlation footprints in that region. In most of the other regions, differences of the order of less than a few percent are seen. {\it Right}: Histogram (expressed as percentage) of the relative differences (in \%) between antenna cross-correlation weights shown on the left binned in intervals of 0.5\%. Over 70\% of the $uv$-cells do not differ by more than 0.5\% and over 90\% of $uv$-cells only differ by $<5$\% for the gridding coarseness used.}} 
  \label{fig:MOFF-FX-uvwts-diff}
\end{figure}

@@

We study the effect of the differences in gridded weights on the image plane. Fig.~\ref{fig:image-psf-diff} shows the difference between the synthesized beams obtained with the two methods. A difference map between the two synthesized beams is shown on the {\bf left}. The amplitude of the difference appears to be modulated by the directional power response of the antenna. {\bf On the right}, in radial bins, the rms of the synthesized beam (gray) and the rms of the difference map (black) are plotted in percentage units relative to the peak (to be read using the axis on the left side of the plot). The antenna power pattern (red; to be read using the scale on the right) is plotted for reference.

@@

The synthesized beam rms is proportional to the antenna power pattern as 
expected from a point spread function uncorrected for the antenna power pattern. 
The rms of differenced synthesized beams is also modulated by the antenna power 
pattern. The rms of the difference is definitely lesser than the rms of the 
synthesized beam in the central regions up to $(l^2+m^2)^{1/2}\lesssim 0.3$. This 
implies that the beams are well matched in the central regions. In the outer 
regions, their mismatch is comparable to the rms of synthesized beams. These findings are consistent with Fig.~\ref{fig:MOFF-FX-psf-diff}. This 
indicates the two synthesized beams are not randomly different from 
each other in which case the rms of the difference would have been 
$\approx \sqrt{2}$ higher than the rms of the each of the synthesized beams. 
This {\bf means} that while differences exist, large fractions of them are still 
well matched to each other even out to the horizon. Thus the rounding errors in 
gridding do not affect the statistics of the images or the synthesized beams.

@@

{\bf One of the advanced features of the MOFF algorithm and its implementation via EPIC is the ability to naturally account for heterogeneity of antennas while still producing images with maximal information \citep{mor11}. Unlike the assumption frequently made in standard radio interferometry that all antennas are identical, EPIC can implement the most generic case where no such assumption can be made. In other words, EPIC does not assume $W_a(\mathbf{r}-\mathbf{r}_a)$ or $B_{ab}(\mathbf{r}-\mathbf{r}_{ab})$ for different antennas are identical. However, any {\it a priori} knowledge available that certain sets of antennas have identical illumination patterns can be easily passed on to EPIC to avoid redundant computations and thus increase efficiency.

@@

This demonstrates that images from EPIC can robustly image data from heterogeneous arrays while being mathematically optimal. Besides being sub-optimal, erroneous assumptions about the homogeneity of an array can lead to significant and systematic mis-estimates (in this case, an underestimate) of the sky model.}

@@

We now consider antenna array layouts described by three essential quantities in radio interferometry, namely, maximum baseline length, number of antennas, and the size of each antenna. Fig.~\ref{fig:parameter-space-bll-nant-instruments} shows the boundaries where the ratio of the number of computations required with visibility-based imaging relative to MOFF imaging is unity. The different colored lines correspond to different antenna sizes (cyan - 1~m$^2$, blue - 7~m$^2$, purple - 16~m$^2$, green - 28~m$^2$, orange - 150~m$^2$, red - 740~m$^2$, gray - 5900~m$^2$). {\bf Dashed lines of each color denote the boundary to the left of which the MOFF algorithm is favoured for the corresponding antenna size and vice versa for visibility-based imaging.}

\begin{figure}
  \includegraphics[width=\columnwidth]{figure12}
  % \includegraphics[width=\columnwidth]{MOFF_FX_crossover_baseline_n-antennas_rho_fov_gridding_legended}
  % \includegraphics[width=\columnwidth]{MOFF_FX_crossover_baseline_n-antennas_rho_fov_gridding_legended.eps}
  \caption{Current and planned instruments in parameter space of baseline length and number of antennas with MOFF and FX. Lines of different colors denote different classes of antenna sizes (cyan - 1~m$^2$, blue - 7~m$^2$, purple - 16~m$^2$, green - 28~m$^2$, orange - 150~m$^2$, red - 740~m$^2$, gray - 5900~m$^2$). Lines of each color denote the boundary to the left of which MOFF algorithm is favoured for the corresponding antenna size. These lines shift rightward with increasing antenna size. The different antennas are color coded by roughly the class of antenna size they fall into. {\bf Thus symbols of one color falling to the left of a line} of the same color indicate MOFF imaging is advantageous for those telescopes and vice versa. For example, MOFF imaging is favoured in HERA-331 and HERA-6769 because they {\bf lie to the left of the orange line} but not so in cases of HERA-19 and HERA-37.}
  \label{fig:parameter-space-bll-nant-instruments}
\end{figure}

{\bf There is an upper limit set by the maximum number of antennas for each antenna size that can be densely packed inside various baseline lengths but is not shown on this figure to avoid crowding. We note that as antenna size increases the maximum number of antennas for a dense packing as a function of baseline length decreases. Hence, this upper limit shifts rightward as antenna size increases. Similarly, with increase in antenna size, $\Ngrid$ also decreases when field of view imaging is achieved with an increased grid spacing equal to antenna size and hence lowers the amount of computations required with the MOFF algorithm. This shifts the dashed lines rightward as well.}
